% LAB 7: Programming Competition
% 
% CSE/IT 107: Intro to Programming in Python
% New Mexico Tech
% 
% Prepared by Christopher Koch
% Fall 2014
\documentclass[11pt]{cselabheader}
% formatting the problem
\newcommand{\cop}[3]{\textbf{#1}~(#2 pts, filename: \texttt{#3})\quad}
\usepackage{IEEEtrantools} 
\theoremstyle{plain}
\newtheorem{thm}{Theorem}
% drawings in latex
\usepackage{tikz}

\fancyhead[R]{Lab 7: Programming Competition}
\title{Lab 7: Programming Competition}

\begin{document}

\maketitle

\hrule

\begin{quotation}
	``First, solve the problem. Then, write the code.''
\end{quotation}
\begin{flushright}
	--- John Johnson
\end{flushright}

\begin{quotation} ``Beware of bugs in the above code; I have only proved it
	correct, not tried it.''
\end{quotation}
\begin{flushright}
	--- Donald Knuth
%\\\textit{Notes on the van Emde Boas construction of priority deques}
%% From corresponde with Peter van Emde Boas about Priority Deques
%% http://www-cs-faculty.stanford.edu/~uno/faq.html
\end{flushright}

\begin{quotation}
	``Openness always deserves recognition.'' (``Offenheit verdient immer
	Anerkennung.'')
\end{quotation}
\begin{flushright}
	--- Otto von Bismarck
% \\\textit{Speeches in the second chamber of Prussian state diet
% (parliament). November 24th, 1849.}

%% Reden in der preußischen Zweiten Kammer, 24. November 1849
%% http://de.wikiquote.org/wiki/Otto_von_Bismarck
%% Translation by Chris
\end{flushright}

\hrule

\section{Introduction}

This lab is solely a problem solving lab.

\pagebreak
\section{Programming Competition}

Please pair up into groups of two or three with somebody in your lab. You will 
enter into the competition together and submit the same files. 

%To submit your answers, find the scoreboard at 
%\begin{center}
%	\vspace{-5mm}
%	\url{http://niedersfeld.cs.nmt.edu/scoreboard/}
%	\vspace{-5mm}
%\end{center}
%Upon submission of your answer, someone will come and check the code to make
%sure you actually found the answer using code and did not use the internet.

To submit your answers, please notify a TA. The whiteboard will serve as the
scoreboard for your lab. The TA will want to see the answer to the ``scoreboard
submission.'' If the TAs are grading another team's submission, do not wait for
them and move on to a new problem.

\textbf{Do not share the problem descriptions or the code with the other lab
section.} Everyone needs to have an equally fair chance of doing well in this
competition.

You may not use the internet to find answers to these problems. Think! If the
concepts are unclear, please get help from a TA.

\subsection{Extra Credit}
This lab will not count for its own grade. Your standing in the competition 
will count for extra credit on your lab final.
The winners of your lab section will get 10\%, the second place will get
9\%, and so on as extra credit on their lab final. 

You have to gain at least 20 points in the competition to qualify for the extra
credit. 

\subsection{Requirements}
Even though this lab ``only'' counts for extra credit, you have to stay the lab
and try your best. \textbf{You will LOSE 10\% on your lab final if you do not
stay and try your best.}

You may choose to do any of the following problems. However, there is no credit
for incomplete or incorrect submissions.

%You may use Google to help find solutions to the problems. However,
%there are two conditions to this: (1) you \textbf{must}
%cite the sources you use in your comments (URLs, etc) and (2) you \textbf{must}
%understand the content you use from these sources. \textbf{Your grader will ask
%you questions to make sure you understand.}

\pagebreak
\subsection{Problems}

\begin{enumerate}[\text{Problem} {[}1{]}~]
	\item \cop{Multiples of 3 and 5}{5}{multiples.py}
		% project euler 1

		If we list all the integers (natural numbers) below 10 that are
		multiples of 3 or 5, we get 3, 5, 6, and 9. The sum of these multiples
		is 23.

		Write a program to find the sum of all the multiples of 3 or 5 below
		1000.

		\textit{Scoreboard submission:} The sum of all multiples of 3 or 5
		below 1000.

	\item \cop{Special Pythagorean Triplet}{10}{pythagoras.py}
		% project euler 9

		A Pythagorean triplet is a set of three integers (natural numbers), $a
		< b < c$ such that $a^2 + b^2 = c^2$.

		For example, $a = 3, b = 4, c = 5$ fits: $3^2 + 4^2 = 9 + 16 = 25 =
		5^2$.

		There exists exactly one Pythagorean triplet for which $a + b + c =
		1000$.

		\textit{Scoreboard submission:} Find the product $abc$ of that special
		Pythagorean triplet.

	\item \cop{Powers}{10}{powers.py}
		% Kaley's idea

		Write a program that determines whether a given integer is a power of
		another integer and what the base and exponent are; for example, $32 = 
		2^5$. Here, $2$ is the base and $5$ is the exponent. You may assume 
		that the integer given is less than $1,000,000$. 

		For example:
		\begin{lstlisting}[style=bash]
A number: 32
32 is 2 to the 5th power.

A number: 25
25 is 5 to the 2nd power.

A number: 23
23 is not a power of any number.
		\end{lstlisting}

		\textit{Scoreboard submission:} Find an integer base and exponent for 
		$823,543$. % 7^7
		% todo: check if there are other possible bases. if there are, make the
		% program look for all of them?

	\item \cop{Fractions}{10}{fractions.py}
		% Source: ACM Freshman Coding Competition, Fall 2013 (Russell White)
		
		Write a program that reads both the numerator and denominator of
		a fraction and displays the reduced fraction. For simplicity, you
		may assume that the numbers entered are less than 10,000. 
		
		For example:
		\begin{lstlisting}[style=bash]
Enter numerator: 4
Enter denominator: 30
Result: 2/15
		\end{lstlisting}

		\begin{lstlisting}[style=bash]
Enter numerator: 25
Enter denominator: 5
Result: 5/1
		\end{lstlisting}

		\textit{Hints}: See Appendix~\ref{sec:gcd} to find out what a GCD is 
		and what Euclid's algorithm is. 

		\textit{Scoreboard submission:} There is none. You have to get this one
		looked at and tested by a TA.

        \item \cop{Crossing Hands}{20}{hands.py}
		% source: http://programmingpraxis.com/2014/02/25/crossing-hands/

		The hands of an analog clock occassionally cross as they revolve around
		the dial. Write a program that determines how many times the hands
		cross in one twelve-hour period, and compute a list of those times
		(down to seconds). Assume you are just using two hands: the hour hand 
		and the minute hand. Also assume that they tick continuously.

		\textit{Scoreboard submission} Add up the hours, minutes, and seconds
		of the times separately and submit it in the format \texttt{H-M-S}, 
		for example \texttt{30-167-421}.

        \item \cop{Finding a Path}{25}{path.py}
          %Source: ENMY Codecomp, Spring 2014 (Tyler Cecil)
          Write a program that finds the shortest valid path between two nodes in a graph.
          Each of the lines of input will be the name of a node (a single, upper-case
          character), followed by a list of the nodes connected to that node. The final
          line of input will a single upper-case letter, designating the destination node.
          You must output a sequence of letters that represents the shortest path to the
          destination. Assume you start at the first given node. You can assume a path will
          always exist. There will be less than 10 nodes given. If multiple valid paths
          exist, any will be accepted.
          
          For example:\\
          \begin{minipage}{.33\linewidth}
            \begin{minipage}{.8\linewidth}
              \begin{lstlisting}[style=bash]
A B D
B C D
C

A B C
              \end{lstlisting}
            \end{minipage}
          \end{minipage}
          \begin{minipage}{.33\linewidth}
            \begin{minipage}{.8\linewidth}
              \begin{lstlisting}[style=bash]
A B C D
B F
D B C
F E
E

A B F E
              \end{lstlisting}
            \end{minipage}
          \end{minipage}
          \begin{minipage}{.33\linewidth}
            \begin{minipage}{.8\linewidth}
              \begin{lstlisting}[style=bash]
A B C D E
B C D E F
C D E F
D E F
E F
F

A B F
              \end{lstlisting}
            \end{minipage}
          \end{minipage}

          \textit{Scoreboard submission:} There is none. You have to get this
          one looked at and tested by a TA.

	\item \cop{Block Areas}{35}{blocks.py}
		% Source: ACM Freshman Coding Competition, Fall 2013 (Russell White)
		% sample solution: http://www.mathopenref.com/coordpolygonarea2.html

		Write a program to take in a series of coordinate points and finds the 
		enclosed area. The points will all be non-negative integers not 
		exceeding 100. All angles in the shape will be 90 degrees. You can 
		assume that no shape will be defined by more than 20 points. 
		
		For example:
		\begin{figure}[!ht]
			\centering
			\begin{tikzpicture}[scale=0.7]
				\node at (2,1) {};% {(2,1)};
				\node at (4,1) {};% {(4,1)};
				\node at (4,3) {};% {(4,3)};
				\node at (3,3) {};% {(3,3)};
				\node at (3,5) {};% {(3,5)};
				\node at (1,5) {};% {(1,5)};
				\node at (1,4) {};% {(1,4)};
				\node at (2,4) {};% {(2,4)};

				\draw (1,5) -- (3,5);
				\draw (1,5) -- (1,4);
				\draw (1,4) -- (2,4);
				\draw (2,4) -- (2,1);
				\draw (2,1) -- (4,1);
				\draw (4,1) -- (4,3);
				\draw (4,3) -- (3,3);
				\draw (3,3) -- (3,5);
			\end{tikzpicture}
			\caption{Example shape}
		\end{figure}

		\begin{lstlisting}[style=bash]
Enter number of points: 8
Enter point: 2 1
Enter point: 4 1
Enter point: 4 3
Enter point: 3 3
Enter point: 3 5
Enter point: 1 5
Enter point: 1 4
Enter point: 2 4

Area: 7
		\end{lstlisting}

		\textit{Scoreboard submission:} Find the area of the polygon described
		by the following 14 coordinates. The order of entering them should not
		matter!
		\[
		(0, 0)~(6, 0)~(6, 1)~(2, 1)~(0, 2)~(2, 2)~(7, 2)
	\]
\[(1, 2)~(1, 3)~(2,
		3)~(2, 4)~(9, 4)~(9, 1)~(7, 1)
		\]

		\begin{figure}[!ht]
			\centering
			\begin{tikzpicture}[scale=0.7]
				\node at (0,0) {};% {(2,1)};
				\node at (6,0) {};% {(4,1)};
				\node at (6,1) {};% {(4,3)};
				\node at (2,1) {};% {(3,3)};
				\node at (2,2) {};% {(3,5)};
				\node at (7,2) {};% {(1,5)};
				\node at (7,1) {};% {(1,4)};
				\node at (9,1) {};% {(2,4)};
				\node at (9,4) {};
				\node at (2,4) {};
				\node at (2,3) {};
				\node at (1,3) {};
				\node at (1,2) {};
				\node at (0,2) {};

				\draw (0,0) -- (6,0);
				\draw (6,0) -- (6,1);
				\draw (6,1) -- (2,1);
				\draw (2,1) -- (2,2);
				\draw (2,2) -- (7,2);
				\draw (7,2) -- (7,1);
				\draw (7,1) -- (9,1);
				\draw (9,1) -- (9,4);
				\draw (9,4) -- (2,4);
				\draw (2,4) -- (2,3);
				\draw (2,3) -- (1,3);
				\draw (1,3) -- (1,2);
				\draw (1,2) -- (0,2);
				\draw (0,2) -- (0,0);
			\end{tikzpicture}
			\caption{Scoreboard submission shape}
		\end{figure}


\end{enumerate}
\pagebreak
%\section{Submission guidelines}
%For submission, you will upload a tar archive containing the following items:
%\begin{itemize}
%	\item Problem files
%	\item A script named \texttt{lab7.script} of you compiling and running 
%		your chosen lab 7 programs (include all cases of operation)
%\end{itemize}

%The name of your tar archive should be
%\begin{center}
%	\vspace{-3mm}
%	\texttt{cse113\_lastname1\_lastname2\_lab7.tar.gz}
%	or
%	\texttt{cse113\_lastname1\_lastname2\_lastname3\_lab7.tar.gz}
%
%	\textbf{Upload your tarball at the end of the lab.}
%\end{center}

\begin{appendices}
	\section{Reducing Fractions -- GCD and Euler's Algorithm}
	\label{sec:gcd}

	Reducing a fraction involves dividing the numerator and denominator by
	the same integers until there is nothing that divides both numerator and
	denominator anymore.

	For example, $\frac{60}{6} = \frac{30}{3} = \frac{10}{1} = 10$. First, I
	divided both numerator and denominator by $2$, and then by $3$. 

	This is where the concept of a GCD comes in. GCD stands for ``greatest
	common divisor'' -- given two numbers such as $60$ and $6$, the GCD is the
	greatest whole number that both numbers are divisible by. Since both $60$ 
	and $6$ are divisible by $6$ and since there is no greater integer than $6$
	that one can divide them both by, the $\gcd(60, 6) = 6$. For another
	example, $\gcd(40, 6) = 2$.

	If you think about it more, it becomes clear that one can reduce a fraction
	by dividing both the numerator and denominator by the GCD of both. In
	essence, given a fraction $\frac{a}{b}$, the reduced fraction would be:
	\[
		\frac{a}{b} = \frac{~\frac{a}{\gcd(a, b)}~}{~\frac{b}{\gcd(a, b)}}
	\]

	How do we find the GCD though? Here, a neat algorithm called Euclid's
	algorithm comes in. Euclid's algorithm takes advantage of the fact that
	when an integer $a$ is divided by another integer $b$, we get a quotient
	$q$ and a remainder $r$. This is called the Division Algorithm (see
	Theorem~\ref{thm:div}). 

	\begin{thm}[Division Algorithm] \label{thm:div}
		Given integers $a$ and $b$, where $b > 0$, there exist unique integers
		$q$ (the quotient) and $r$ (the remainder) such that 
		\[ a = bq + r\text{, where }0 \leq r < b. \]
	\end{thm}

	For example, when $5$ is divided by $2$, we get a quotient of $2$ and a
	remainder of $1$, since $5 = 2 \times 2 + 1$. You have already taken 
	advantage of this algorithm when you validated credit card numbers. 

	In C, we can find the quotient by doing integer division and the remainder
	using the modulus operator \%.

	\subsection{Euclid's Algorithm}

	Euclid's algorithm finds the GCD by repeatedly using the division
	algorithm. Given integers $a$ and $b$, where $a > b$, that we want to find 
	the GCD of, we
	find the quotient $q$ and $r$. If the remainder is $0$, $b$ is the GCD. If
	the remainder is not $0$, we repeat the same steps, but we let $a = b$ and
	$b = r$. 

	In essence, it goes:
	\begin{IEEEeqnarray*}{RCLCL}
		a &=& bq + r &\text{ where }&0 \leq r < b \\
		b &=& rq_2 + r_2 &\text{ where }&0 \leq r_2 < r \\
		r &=& r_2q_3 + r_3 &\text{ where }&0 \leq r_3 < r_2
	\end{IEEEeqnarray*}
	So we are saying that $\gcd(a, b) = \gcd(b, r) = \gcd(r, r_2) = \cdots =
	\gcd(r_n, 0) = r_n$. Notice that the quotient is not used at all, only the
	given integers $a, b$ and the remainder $r$ matter.
	
	Notice that $r_n < \cdots < r_3 < r_2 < r < b$.

	It seems natural to write this with recursion. In pseudocode, with recursion:
	\begin{lstlisting}
euclids(a, b) 
	if(a < b) 
		return euclids(b, a)
	
	r = a modulus b
	if(r == 0)
		return b
	else
		return euclids(b, r)
	\end{lstlisting}

\end{appendices}
\end{document}
